Accesible, realistic genome simulation with selection using \stdpopsim

### 2025-03-10 analysis2_manuscript.tex
up to discussion

\section*{Estimation of the Distribution of Fitness effects}
    \label{dfe}
    Another common application of population genetic inference is to estimate the distribution of fitness effects (DFE) of new mutations
    from genetic data. The new framework of modeling selection in \stdpopsim makes it an ideal tool for easily benchmarking and comparing
    the performance of different methods for inferring the DFE. 
    %IG: I don't think we need this intro.
    %We performed simulations using the 
    %human and vaquita porpoise models described above, with complex demographic histories and models of selection acting on exons.
    %
    A challenge when comparing different inferred and simulated distributions of the selection coefficient ($s$) is differing convention regarding its definition.
    The simulator SLiM and the inference tool \grapes define the selection coefficient $s$ such that a homozygote has fitness $1+s$.
    On the other hand, \dadicli and \polydfe define the selection coefficient such that a homozygote has fitness $1+ 2s$.
     %https://academic.oup.com/genetics/article/207/3/1103/5930676
    Moreover, inferred distributions of selection coefficients are typically scaled by an inferred ancestral
    effective population size $N_e$, and different methods assume different scaling.
    \dadicli infers the distribution of $2 N_e s$ while \polydfe and \grapes infer the distribution of $4 N_e s$.
    %https://academic.oup.com/genetics/article/207/3/1103/5930676, https://academic.oup.com/genetics/article/216/2/559/6066183
    For comparison, here we have normalized all inference results to the convention of SLiM (see \textbf{Methods}).
    % Add citation to fastDFE as the successor to polyDFE: https://academic.oup.com/mbe/article/41/5/msae070/7641109 ?

@@

    We start by examining DFE inference from human genomes simulated under an out-of-Africa demographic model
    with selection acting on exons (see \textbf{Methods}).
    In this setting, the DFE is inferred separately for each of the three simulated populations
    using three different DFE inference methods: \dadicli \citep{Huang2023}, \polydfe \citep{tataru2020polydfe}, 
    and \grapes \citep{galtier2016adaptive}.
    In each case we note that the DFE is estimated from segregating sites only,
    without the use of substitution data, as \stdpopsim's models currently do not
    contain outgroup populations or species that can be used to estimate divergence.
    %IG: no need to map figures. Just describe results and cite figs
    %Figure \ref{fig:homsap-dfe.constant} shows estimates from a single, constant size population of humans, 
    %while Figure \ref{fig:homsap-dfe.ooa} shows estimates from a model of human out-of-Africa demography.
    %In the out-of-Africa model (Fig.~\ref{fig:homsap-dfe.ooa}) the performance of the methods is more mixed. 
    Overall, inferred DFEs varied across populations and inference methods (Figure \ref{fig:homsap-dfe.ooa}).
    Note that while the estimation accuracy for the DFE parameters themselves (Figure \ref{fig:homsap-dfe.ooa}A-B) is useful,
    examining the difference between the actual and inferred distributions can be more helpful for showing how well a method
    has recovered the true DFE (Figure \ref{fig:homsap-dfe.ooa}C).
    The distributions of $s$ inferred by \polydfe and \dadicli had higher mean values (in absolute value)
    but smaller shape parameters, while \grapes inferred distributions with lower mean values and shape parameters.
    As a result, \polydfe and \dadicli over-estimate the fraction of mutations under strong selection ($s<-0.1$),
    while \grapes considerably over-estimates the fraction of mutations under weak selection ($s\geq-0.0001$).
    % IG: commented this out to simplify the basic observation
    % and under-estimates the fraction of mutations with $s<-0.01$.

@@

    %IG: describe experiment and cite figure later
    %In Figures \ref{fig:vaquita-dfe.constant} and \ref{fig:vaquita-dfe} we show a comparison of estimates
    Next, we benchmarked these three DFE-inference methods on simulations of the vaquita porpoise genome.
    Recall that these simulations apply a gamma-distributed selection coefficient to mutations in exons
    and a relationship between the selection and dominance coefficients (see \textbf{Methods}).
    In this setting, all methods perform uniformly worse than in the human genome simulations,
    with consistent underestimation of the mean selection coefficient    
    and overestimation of the shape parameter (Figure \ref{fig:vaquita-dfe}A-B). 
    Similar patterns are observed when re-running the simulations with a simpler demographic model
    without population size changes (Figure \ref{fig:vaquita-dfe.constant}).
    We suspect that this is due to the fact that all DFE inference methods used here assume an additive model ($h = \frac{1}{2}$),
    while mutations in the simulated genomes were more recessive ($h < \frac{1}{2}$), and their dominance decreased with the strength of selection. 
    To examine the influence of this model misspecification, we compared the inferred distributions of $s$ to the 
    simulated distribution of $2hs$, which would have been equal to $s$ under an additive model (Figure \ref{fig:vaquita-dfe}C).
    Reassuringly, all three methods infer distributions that appear to fit the main mode of the simulated distribution of $2hs$.
    This is likely because deleterious alleles are typically at low frequency and thus heterozygous, 
    where their selective effect is $h s$ rather than $s$.
    %KEL: I moved the last sentence from here to the Discussion and fleshed it out

@@

    We evaluated the statistical power to detect selective sweeps for three methods:
    1) \sweepfinder \citep{degiorgio2016sweepfinder2},
    2) \diploshic \citep{kern2018diplos},
    and 3) reduced diversity ($\pi$).
    For each method, we determined a threshold for its test-statistic 
    using each of the two null models separately (neutral and BGS) and controlling 
    for $5\%$ false positive rate.
    We then applied each method to data simulated with sweeps for two sampled populations (YRI and CEU)
    in each of the 100 locations along human chromosome 1 (Figure \ref{fig:chr1_power}).
    First, we notice that power to detect sweeps is slightly lower when the test statistic threshold
    is determined using simulated data with background selection (BGS null model).
    %This observation is consistent with what was previously shown by \cite{schrider2020background},
    %but the effect is relatively minor. DRS: cutting this because my paper showed that power decreases when our test set has BGS and our training set does not, so not quite the same.
    This is likely because the strength of selection acting on exons is relatively weak, and the
    sweep detection methods we have used are relatively robust to the effects of background selection. 
    Second, the power to detect sweeps is generally higher in the YRI sample
    than in the CEU sample, which is consistent with the fact that YRI has a larger ancestral effective population size and
    thus both stronger selection and more variation to detect a sweep \citep[e.g.,][]{simonsen1995properties}.
    Further, the CEU population experienced the out-of-Africa bottleneck
    that may mute the signal of a sweep.  
    Finally and most importantly, the power to detect sweeps varies considerably along the chromosome,
    and this variation is much greater than the variation in power between CEU and YRI or between
    null models. This is likely due to the effects of recombination rate heterogeniety (Figure \ref{fig:chr1_power}).% as well
    %as the density of functional elements along the chromosome.
    %DRS: cut the density of func elements because that stuff doesn't seem to matter much here

    We compared the overall performance of the three sweep-detection methods using
    receiver operating characteristic (ROC) curves (Figure \ref{fig:roc-curves}).
    Different methods perform better for each of the sampled populations,
    with \sweepfinder performing better than the other two on the CEU sample,
    and \diploshic performing better on the YRI sample.
    This is may result from the fact that we trained \diploshic on simulations under a constant size population,
    which more closely matches the demographic history of YRI, than that of CEU, which 
    is characterized by a strong recent bottleneck followed by population expansion.
    \drscomment{note: I am couching the language above a bit because we have previously shown that shic doesn't suffer in the sweep vs. nonsweep question when doing other train-on-constant-test-on-nonequilib benchmarks. Could be something else going on here (or some interaction between demog misspec and rec rate variation), but in any case I am not certain of the cause so I am toning down our certainty slightly.}
    Interestingly, the simple method that uses reduced diversity ($\pi$) as an indicator
    for a selective sweep performs reasonably well overall (especially on the CEU sample).
    However, its accuracy is much more variable along the chromosome compared to the other two method (see Figure \ref{fig:chr1_power}),
    and for quite a few locations along the chromosome it has power below $50\%$
    when applied to both sample sets.
    In that sense, \sweepfinder appears to have the least variation in power along the chromosome or across populations.

@@

    To further dissect the reasons for high variance along the chromosome in power to detect sweeps,
    we examined the relationship between power and local recombination rate for our sweep simulations ( Figure \ref{fig:power-recomb}).
    We see that power to detect sweeps is
    a decreasing function of local recombination rate for \sweepfinder and reduced diversity,
    which is consistent with sweeps having a larger genomic footprint in regions of low
    recombination.
    \diploshic on the other hand shows a slight increase in power with recombination rates,
    probably because it is designed to distinguish regions where a sweep happened from regions linked to a sweep,
    and if the recombination rate is too low (or selection too strong) then the window examined by \diploshic may
    not be large enough to exhibit enough of a recovery in patterns of diversity flanking the sweep.
    These results confirm that reduced genetic diversity performs well as an indicator for selective sweeps only in regions with
    low recombination rate.
    The other two methods, which were specifically designed for sweep detection, are considerably more robust 
    to the effects of recombination rate than the simple indicator of reduced genetic diversity.
    These methods are particularly robust to recombination rate variation when applied to 
    the YRI sample, especially when using a neutral null model to calibrate the test statistics.
    The other null model, which assumes background selection from exons,
    appears to consistently lead to reduced power for detecting sweeps regardless of the recombination rate.
    Interestingly, we do not see a clear correlation between power to detect sweeps and
    local exon density (Figure \ref{fig:power-exon}).
    This is somewhat surprising given that background selection is expected to somewhat interfere with the sweep signature.
    This might suggest that the effects of realistic background selection on genome variation
    may not be as similar to those of sweeps as previously thought \citep{schrider2020background}.

>>>>

\section*{Estimation of the Distribution of Fitness effects}
    \label{dfe}
    Another common application of population genetic inference is to estimate the distribution of fitness effects (DFE) of new mutations
    from genetic data. The new framework of modeling selection in \stdpopsim makes it an ideal tool for easily benchmarking and comparing
    the performance of different methods for inferring the DFE.

    A challenge when comparing different inferred and simulated distributions of the selection coefficient ($s$)
    is differing conventions in the literature regarding its definition.
    The simulator \texttt{SLiM} and the inference tool \grapes define the selection coefficient $s$ such that a homozygote has fitness $1+s$.
    On the other hand, \dadicli and \polydfe define the selection coefficient such that a homozygote has fitness $1+ 2s$.
     %https://academic.oup.com/genetics/article/207/3/1103/5930676
    Moreover, inferred distributions of selection coefficients are typically scaled by an inferred ancestral
    effective population size $N_e$, and different methods assume different scaling.
    \dadicli infers the distribution of $2 N_e s$ while \polydfe and \grapes infer the distribution of $4 N_e s$.
    %https://academic.oup.com/genetics/article/207/3/1103/5930676, https://academic.oup.com/genetics/article/216/2/559/6066183
    For comparison, here we have normalized all inference results to the convention of \texttt{SLiM} (see \textbf{Methods}).
    % Add citation to fastDFE as the successor to polyDFE: https://academic.oup.com/mbe/article/41/5/msae070/7641109 ?

@@

    We start by examining DFE inference from human genomes simulated under an out-of-Africa demographic model
    with selection acting on exons (see \textbf{Methods}).
    In this setting, the DFE is inferred separately for each of the three simulated populations
    using three different DFE inference methods: \dadicli \citep{Huang2023}, \polydfe \citep{tataru2020polydfe}, 
    and \grapes \citep{galtier2016adaptive}.
    In each case we note that the DFE is estimated from segregating sites only,
    without the use of substitution data, as \stdpopsim's models currently do not
    contain outgroup populations or species that can be used to estimate divergence.

    Overall, inferred DFEs varied across populations and inference methods (Figure \ref{fig:homsap-dfe.ooa}).
    Note that while the estimation accuracy for the DFE parameters themselves (Figure \ref{fig:homsap-dfe.ooa}A-B) is useful,
    examining the difference between the actual and inferred distributions can be more helpful for showing how well a method
    has recovered the true DFE (Figure \ref{fig:homsap-dfe.ooa}C).
    The distributions of $s$ inferred by \polydfe and \dadicli had higher mean values (in absolute value)
    but smaller shape parameters, while \grapes inferred distributions with lower mean values and shape parameters.
    As a result, \polydfe and \dadicli over-estimate the fraction of mutations under strong selection ($s<-0.1$),
    while \grapes considerably over-estimates the fraction of mutations under weak selection ($s\geq-0.0001$).
    % IG: commented this out to simplify the basic observation
    % and under-estimates the fraction of mutations with $s<-0.01$.

@@

    Next, we benchmarked these three DFE-inference methods on simulations of the vaquita porpoise genome.
    Recall that these simulations apply a gamma-distributed selection coefficient to mutations in exons
    with a relationship between the selection and dominance coefficients (see \textbf{Methods}).
    In this setting, all methods perform uniformly worse than in the human genome simulations,
    with consistent underestimation of the mean selection coefficient    
    and overestimation of the shape parameter (Figure \ref{fig:vaquita-dfe}A-B). 
    Similar patterns are observed when re-running the simulations with a simpler demographic model
    without population size changes (Figure \ref{fig:vaquita-dfe.constant}).
    We suspect that this is due to the fact that all DFE inference methods used here assume an additive model ($h = \frac{1}{2}$),
    while mutations in the simulated genomes were more recessive ($h < \frac{1}{2}$), and their dominance decreased with the strength of selection. 
    To examine the influence of this model misspecification, we compared the inferred distributions of $s$ to the 
    simulated distribution of $2hs$, which would have been equal to $s$ under an additive model (Figure \ref{fig:vaquita-dfe}C).
    Reassuringly, all three methods infer distributions that appear to fit the main mode of the simulated distribution of $2hs$.
    This is likely because deleterious alleles are typically at low frequency and thus heterozygous, 
    where their selective effect is $h s$ rather than $s$.

@@

    We evaluated the statistical power to detect selective sweeps for three methods:
    1) \sweepfinder \citep{degiorgio2016sweepfinder2},
    2) \diploshic \citep{kern2018diplos},
    and 3) reduced diversity ($\pi$).
    For each method, we determined a threshold for its test-statistic 
    using each of the two null models separately (neutral and BGS) and controlling 
    for a $5\%$ false positive rate.
    We then applied each method to data simulated with sweeps for two samples (YRI and CEU)
    in each of 100 locations along human chromosome 1 (Figure \ref{fig:chr1_power}).
    First, we notice that power to detect sweeps is slightly lower when the test statistic threshold
    is determined using simulated data with background selection (BGS null model).
    %This observation is consistent with what was previously shown by \cite{schrider2020background},
    %but the effect is relatively minor. DRS: cutting this because my paper showed that power decreases when our test set has BGS and our training set does not, so not quite the same.
    This is likely because the strength of selection acting on exons is relatively weak, and the
    sweep detection methods we have used are relatively robust to the effects of background selection. 
    Second, the power to detect sweeps is generally higher in the YRI sample
    than in the CEU sample, which is consistent with the fact that YRI has a larger ancestral effective population size and
    thus both stronger selection and more variation to detect a sweep \citep[e.g.,][]{simonsen1995properties}.
    Conversely, the CEU population experienced the out-of-Africa bottleneck that may mute the signal of a sweep.  
    Finally and most importantly, the power to detect sweeps varies considerably along the chromosome,
    and this variation is much greater than the variation in power between CEU and YRI or between
    null models. This is likely due to the effects of recombination rate heterogeniety (Figure \ref{fig:chr1_power}).% as well
    %as the density of functional elements along the chromosome.
    %DRS: cut the density of func elements because that stuff doesn't seem to matter much here

    We compared the overall performance of the three sweep-detection methods using
    receiver operating characteristic (ROC) curves (Figure \ref{fig:roc-curves}).
    Different methods perform better for each of the sampled populations,
    with \sweepfinder performing better than the other two on the CEU sample,
    and \diploshic performing better on the YRI sample.
    This is may result from the fact that we trained \diploshic on simulations under a constant size population,
    which more closely matches the demographic history of YRI, than that of CEU, which 
    is characterized by a strong recent bottleneck followed by population expansion.
    %\drscomment{note: I am couching the language above a bit because we have previously shown that shic doesn't suffer in the sweep vs. nonsweep question when doing other train-on-constant-test-on-nonequilib benchmarks. Could be something else going on here (or some interaction between demog misspec and rec rate variation), but in any case I am not certain of the cause so I am toning down our certainty slightly.}
    Interestingly, the simple method that uses reduced diversity ($\pi$) as an indicator
    for a selective sweep performs reasonably well overall (especially on the CEU sample).
    However, its accuracy is much more variable along the chromosome compared to the other two method (see Figure \ref{fig:chr1_power}),
    and for quite a few locations along the chromosome it has power below $50\%$
    when applied to both sample sets.
    In that sense, \sweepfinder appears to have the least variation in power along the chromosome or across populations.

@@

    To further dissect the reasons for high variance along the chromosome in power to detect sweeps,
    we examined the relationship between power and local recombination rate for our sweep simulations ( Figure \ref{fig:power-recomb}).
    We see that power to detect sweeps is
    a decreasing function of local recombination rate for \sweepfinder and reduced diversity,
    which is consistent with sweeps having a larger genomic footprint in regions of low
    recombination.
    \diploshic on the other hand shows a slight increase in power with recombination rates,
    presumably because it is designed to distinguish regions where a sweep happened from regions linked to a sweep,
    and if the recombination rate is too low (or selection too strong) then the genomic window examined by \diploshic may
    not be large enough to exhibit enough of a recovery in patterns of diversity flanking the sweep.
    These results confirm that reduced genetic diversity performs well as an indicator for selective sweeps only in regions with
    low recombination rate.
    The other two methods, which were specifically designed for sweep detection, are considerably more robust 
    to the effects of recombination rate than the simple indicator of reduced genetic diversity.
    These methods are particularly robust to recombination rate variation when applied to 
    the YRI sample, especially when using a neutral null model to calibrate the test statistics.
    The other null model, which assumes background selection from exons,
    appears to consistently lead to reduced power for detecting sweeps regardless of the recombination rate.
    Interestingly, we do not see a clear correlation between power to detect sweeps and
    local exon density (Figure \ref{fig:power-exon}).
    This is somewhat surprising given that background selection is expected to somewhat interfere with the sweep signature.
    This might suggest that the effects of realistic background selection on genome variation
    may not be as similar to those of sweeps as previously thought \citep{schrider2020background}.
